### 2025-07-10 paper/quarto_aaai/paper.tex
Trim to 7 pages

\iffalse
%Example, Multiple Authors, ->> remove \iffalse,\fi and place them surrounding AAAI title to use it
% \title{My Publication Title --- Multiple Authors}
\author {
    % Authors
    First Author Name\textsuperscript{\rm 1},
    Second Author Name\textsuperscript{\rm 2},
    Third Author Name\textsuperscript{\rm 1}
}
\affiliations {
    % Affiliations
    \textsuperscript{\rm 1}Affiliation 1\\
    \textsuperscript{\rm 2}Affiliation 2\\
    firstAuthor@affiliation1.com, secondAuthor@affilation2.com, thirdAuthor@affiliation1.com
}
\fi

% REMOVE THIS: bibentry
% This is only needed to show inline citations in the guidelines document. You should not need it and can safely delete it.
\usepackage{bibentry}
% END REMOVE bibentry
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amsthm}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\usepackage{placeins}
\usepackage{siunitx}
\sisetup{uncertainty-mode = separate}

@@

Part of the problem is that the high degrees of freedom provide room for
many solutions that are locally optimal with respect to narrow
objectives \citep{wilson2020case}.\footnote{We follow the standard ML
  convention, where ``degrees of freedom'' refer to the number of
  parameters estimated from data.} As one example, research on the
``lottery ticket hypothesis'' suggests that modern neural networks can
be pruned by up to 90\% without losing predictive performance
\citep{frankle2018lottery}. Thus, looking at the predictive performance
alone, found solutions may seem to provide compelling explanations for
the data, when in fact they are based on purely associative and
semantically meaningless patterns. This poses two related challenges.
Firstly, there is no dependable way to verify if learned representations
correspond to meaningful, plausible explanations. Secondly, even if we
resolve this challenge, it remains undecided how to ensure that machine
learning models can \emph{only} learn valuable explanations.

@@

\noindent From a Bayesian perspective, setting
\(\mathbf{x}^+[d] := \mathbf{x}^\prime[d]\) can be understood as
assuming a point mass prior for \(p(\mathbf{x}^+)\) with respect to
feature \(d\). Intuitively, we think of this as ignoring implausibility
costs with respect to immutable features, which effectively forces the
model to instead seek plausibility with respect to the remaining
features. This can be expected to result in lower overall sensitivity to
immutable features, which we investigate empirically in
Section~\ref{sec-experiments}. Under certain conditions, this result
holds theoretically:\footnote{For the proof, see the supplementary
  appendix.}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  To what extent does the CT objective in Equation 1 induce models to
  learn plausible explanations?
\item
  To what extent does the CT objective produce more favorable
  algorithmic recourse outcomes in the presence of actionability
  constraints?
\item
  What are the effects of hyperparameter selection wrt. the CT
  objective?
\end{enumerate}

@@

Our key outcome of interest is improvement in explainability (Def.
\ref{def-explainability}). Thus, we focus primarily on the plausibility
and cost of faithfully generated counterfactuals at test
time\footnote{Additional metrics such as validity and redundancy are
  reported in the appendix.}. To measure the cost, we follow the
standard convention of using distances (\(\ell_1\)-norm) between
factuals and counterfactuals as a proxy. For plausibility, we assess how
similar CEs are to the observed samples in the target domain,
\(\mathbf{X}^+\subset\mathcal{X}^+\). We rely on the metric used by
\citet{altmeyer2024faithful},
\begin{equation}\phantomsection\label{eq-impl-dist}{
\text{IP}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
}\end{equation} and introduce a novel divergence metric,
\begin{equation}\phantomsection\label{eq-impl-div}{
\text{IP}^*(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
}\end{equation} where \(\mathbf{X}^\prime\) denotes a collection of
counterfactuals and \(\text{MMD}(\cdot)\) is the unbiased estimate of
the squared population maximum mean discrepancy, proposed by
\citet{gretton2012kernel}. The metric in Equation~\ref{eq-impl-div} is
equal to zero if and only if the two distributions are exactly the same,
\(\mathbf{X}^\prime=\mathbf{X}^+\).

We also assess the predictive performance of models using standard
metrics, including robust accuracy estimated on adversarially perturbed
data using FGSM \citep{goodfellow2014explaining}.

We run experiments with three gradient-based generators: \emph{Generic}
of \citet{wachter2017counterfactual} as a simple baseline approach,
\emph{REVISE} \citep{joshi2019realistic} that aims to generate plausible
counterfactuals using a surrogate Variational Autoencoder (VAE), and
\emph{ECCo}---the generator of \citet{altmeyer2024faithful} but without
the conformal prediction component---as a method that directly targets
both faithfulness and plausibility of the counterfactuals.

We make use of nine classification datasets common in the CE/AR
literature. Four of them are synthetic with two classes and different
characteristics: linearly separable clusters (\emph{LS}), overlapping
clusters (\emph{OL}), concentric circles (\emph{Circ}), and interlocking
moons (\emph{Moon}). These datasets are generated using the library of
\citet{altmeyer2023explaining} and we present them in the supplementary
appendix. Next, we have four real-world binary tabular datasets from the
domain of economics: \emph{Adult} (a.k.a. Census data) of
\citet{becker1996adult2}, California housing (\emph{CH}) of
\citet{pace1997sparse}, Default of Credit Card Clients (\emph{Cred}) of
\citet{yeh2016default}, and Give Me Some Credit (\emph{GMSC}) from
\citet{kaggle2011give2}. Finally, for the convenience of illustration,
we use the 10-class \emph{MNIST} dataset \citep{lecun1998mnist}.

@@

Our main quantitative results for the \emph{MLP} are summarised in
Table~\ref{tbl-main}, which shows average outcomes along with
bootstrapped standard errors for key metrics. The following example
illustrates how to read Table~\ref{tbl-main}.

@@

\subsubsection{Synthetic Example: Prediction of Credit
Defaults}\label{synthetic-example-prediction-of-credit-defaults}

Figure~\ref{fig-poc} presents results for a linear classifier fitted to
\emph{LS}, which complies with the data assumptions in Proposition
\ref{prp-mtblty}. The four panels show the outcomes for different
training procedures: in panels (a) and (c) we have trained the models
conventionally, while in panels (b) and (d) we have applied CT. For
illustrative purposes, suppose the first feature represents \emph{debt}
(mutable) and the second feature represents \emph{age} (immutable) of
loan applicants seeking counterfactual explanations for moving to the
target class: loan provided (orange).

@@

In panel (c) of Figure~\ref{fig-poc}, the CEs involve substantial
reductions in \emph{debt} for younger applicants. By comparison,
counterfactual paths are shorter on average in panel (d) where we have
protected the immutable \emph{age} as described in
Section~\ref{sec-constraints}. Due to the classifier's lower sensitivity
to \emph{age}, recommendations with respect to \emph{debt} are much more
homogenous and do not unfairly punish younger individuals. These
counterfactuals are also plausible with respect to the mutable feature,
despite requiring smaller debt reductions on average, resulting in
smaller costs to individuals. This result is quantified for the
non-linear case in column three of Table~\ref{tbl-main}, which shows the
\(\%\)-reduction in costs averaged across valid counterfactuals. Once
again, the impact of CT is statistically significant and substantial
(14\%). Thus, we consider the model in panel (d) as the most explainable
according to Def. \ref{def-explainability}. In the following, we present
the results for all remaining datasets.

@@

We find that CT generally leads to substantial and statistically
significant improvements in plausibility: average reductions in \(IP\)
range from around 7\% for \emph{MNIST} to almost 60\% for \emph{Circ};
for the real-world tabular datasets they are around 12\% for both
\emph{CH} and \emph{Cred} and almost 25\% for \emph{GMSC}; for
\emph{Adult} and \emph{OL} we find no significant impact of CT on
\(IP\). Reductions in \(IP^{*}\) are even more substantial and generally
statistically significant, although the average degree of uncertainty is
higher than for \(IP\): average reductions range from around 20\%
(\emph{Moons}) to almost 90\% (\emph{Circ}). The only negative findings
for \emph{OL} and \emph{MNIST} are statistically insignificant and, in
the latter case, do not align with qualitative findings, which are much
more plausible for CT (Figure~\ref{fig-mnist}).

@@

\caption{\label{fig-mnist}Explanations for \emph{MNIST} for BL (top) and
CT (bottom). First column is a random factual 0 (blue). Columns 2 to 5
are corresponding \emph{ECCo} counterfactuals in target classes 1 to 4.
Columns 6 to 10 show integrated gradients averaged over all test images
in classes 5 to 9.}

@@

We also find that CT can reduce sensitivity to immutable, protected
features and thus lead to less costly counterfactual outcomes as
illustrated in Figure~\ref{fig-poc}. In column three of
Table~\ref{tbl-main}, we impose mutability constraints on selected
features and compute the reduction in average costs of counterfactuals
associated with CT compared to the weak baseline: for synthetic
datasets, we always protect the first feature; for all real-world
tabular datasets we could identify and protect an \emph{age} variable;
for \emph{MNIST}, we protect the five top and bottom rows of digits.
Reductions in costs are overwhelmingly positive and significant of up to
nearly 60\% for \emph{GMSC}. While the estimated cost reductions for
\emph{Adult} and \emph{MNIST} are not significant,
Figure~\ref{fig-mnist} (columns 6-10) demonstrates that CT does have the
expected effect: sensitivity to protected features as per integrated
gradients is drastically reduced\footnote{Additional findings for
  integrated gradients are reported in the appendix.}. In the case of
\emph{Cred}, average costs increase, likely because any potential
benefits from protecting the \emph{age} are outweighed by the increase
in costs required for greater plausibility.

@@

\caption{\label{tbl-main}Key performance metrics and bootstrapped
standard errors for all datasets. \textbf{Plausibility} (columns 1-2):
percentage reduction in implausibility for \(IP\) and \(IP^*\),
respectively; \textbf{Actionability} (column 3): percentage reduction in
costs with protected features. \textbf{Accuracy} (columns 4-7): test
accuracies and robust accuracies (\(\text{Acc}^*\)) for CT and the
baseline (BL). Counterfactual outcomes in columns 1-3 are aggregated
across bootstrap samples and varying degrees of the energy penalty
\(\lambda_{\text{egy}}\) used for \emph{ECCo} at test time. Standard
errors for accuracy are bootstrapped from the test set.}

@@

\emph{CT increases the training time of models.} Just like adversarial
training, CT is more resource-intensive than conventional training
regimes. Higher numbers of CEs per iteration improve the quality of
learned representations, but they also increase the number of
computations. As our codebase is not performance optimized, grids of 270
settings took up to four hours for more demanding datasets in our
experiments, when ran on 34 2GB CPUs (see supplementary appendix). Other
than optimization, three factors mitigate this effect: (1) CT yields
itself to parallel execution; (2) it amortizes the cost of CEs for the
training samples; (3) it can be used to fine-tune conventionally-trained
models.

We also highlight several important directions for future research.
Firstly, it is an interesting challenge to extend CT beyond
classification settings. Our formulation relies on the distinction
between non-target class(es) \(y^{-}\) and target class(es) \(y^{+}\) to
generate counterfactuals through Equation~\ref{eq-obj}. While \(y^{-}\)
and \(y^{+}\) can be arbitrarily defined, CT requires the output space
\(\mathcal{Y}\) to be discrete. Thus, it does not apply to ML tasks
where the change in outcome cannot be readily discretized. Focus on
classification is a common choice in research on CEs and AR. Other
settings have attracted some interest, e.g., regression
\citep{spooner2021counterfactual}, but there is little consensus how to
robustly extend the notion of CEs.\\
Secondly, CT is susceptible to training instabilities. This problem has
been recognized for JEMs \citep{grathwohl2020your} and even though we
depart from the SGLD sampling, we still encounter variability in
outcomes. CT is exposed to two potential sources of instabilities: (1)
the energy-based contrastive divergence term, \(\text{div}(\cdot)\), in
Equation~\ref{eq-div}, and (2) the underlying explainers. We find
several promising ways to mitigate this problem: regularizing energy
(\(\lambda_{\text{reg}}\)), generating sufficiently many counterfactuals
during each epoch, and including only mature counterfactuals in
\(\text{div}(\cdot)\).\\
Finally, we believe that it is possible to substantially improve
hyperparameter selection procedures, and thus performance. In this work,
we have relied exclusively on grid searches for this task. Future work
on CT could benefit from investigating more sophisticated approaches.
Notably, our method is iterative which makes methods such as Bayesian or
gradient-based optimization applicable \citep{bischl2023hyperparameter}.

To conclude, state-of-the-art machine learning models are prone to
learning complex representations that cannot be interpreted by humans.
Existing explainability solutions cannot guarantee that explanations
agree with the model's learned representation of data. As a step towards
addressing this challenge, we introduced counterfactual training, a
novel training regime that incentivizes highly-explainable models. Our
approach leads to explanations that are both plausible (compliant with
the underlying data-generating process) and actionable (compliant with
user-specified mutability constraints), and thus meaningful to their
recipients. Through extensive experiments, we demonstrated that CT
satisfies its objective while promoting robustness and preserving the
predictive performance of models. It can also be used to fine-tune
conventionally-trained models and achieve similar gains. Thus, our work
showcases that it is practical to simultaneously improve models and
their explanations.

>>>>

\iffalse
%Example, Multiple Authors, ->> remove \iffalse,\fi and place them surrounding AAAI title to use it
% \title{My Publication Title --- Multiple Authors}
\author {
    % Authors
    First Author Name\textsuperscript{\rm 1},
    Second Author Name\textsuperscript{\rm 2},
    Third Author Name\textsuperscript{\rm 1}
}
\affiliations {
    % Affiliations
    \textsuperscript{\rm 1}Affiliation 1\\
    \textsuperscript{\rm 2}Affiliation 2\\
    firstAuthor@affiliation1.com, secondAuthor@affilation2.com, thirdAuthor@affiliation1.com
}
\fi
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amsthm}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\usepackage{placeins}
\usepackage{siunitx}
\sisetup{uncertainty-mode = separate}
\DeclareMathSizes{10}{9}{7}{6.5}

@@

Part of the problem is that the high degrees of freedom---high number of
parameters estimated from data---provide room for many solutions that
are locally optimal with respect to narrow objectives
\citep{wilson2020case}. As one example, research on the ``lottery ticket
hypothesis'' suggests that modern neural networks can be pruned by up to
90\% without losing predictive performance \citep{frankle2018lottery}.
Thus, looking at the predictive performance alone, found solutions may
seem to provide compelling explanations for the data, when in fact they
are based on purely associative and semantically meaningless patterns.
This poses two related challenges. Firstly, there is no dependable way
to verify if learned representations correspond to meaningful, plausible
explanations. Secondly, even if we resolve this challenge, it remains
undecided how to ensure that machine learning models can \emph{only}
learn valuable explanations.

@@

\noindent From a Bayesian perspective, setting
\(\mathbf{x}^+[d] := \mathbf{x}^\prime[d]\) can be understood as
assuming a point mass prior for \(p(\mathbf{x}^+)\) wrt. feature \(d\).
Intuitively, we think of this as ignoring implausibility costs of
immutable features, which effectively forces the model to instead seek
plausibility through the remaining features. This can be expected to
result in lower overall sensitivity to immutable features, which we
investigate empirically in Section~\ref{sec-experiments}. Under certain
conditions, this result holds theoretically; for the proof, see the
supplementary appendix:

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  To what extent does the CT objective in Equation 1 induce models to
  learn plausible explanations?
\item
  To what extent does CT lead to more favorable AR outcomes in the
  presence of actionability constraints?
\item
  What are the effects of hyperparameter selection on CT?
\end{enumerate}

@@

Our focus is the improvement in explainability (Def.
\ref{def-explainability}). Thus, we primarily look at the plausibility
and cost of faithfully generated counterfactuals at test time. Other
metrics, such as validity and redundancy, are reported in the
supplementary appendix. To measure the cost, we follow the standard
proxy of distances (\(\ell_1\)-norm) between factuals and
counterfactuals. For plausibility, we assess how similar CEs are to the
observed samples in the target domain,
\(\mathbf{X}^+\subset\mathcal{X}^+\). We rely on the metric used by
\citet{altmeyer2024faithful},
\begin{equation}\phantomsection\label{eq-impl-dist}{
\text{IP}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
}\end{equation} and introduce a novel divergence metric,
\begin{equation}\phantomsection\label{eq-impl-div}{
\text{IP}^*(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
}\end{equation} where \(\mathbf{X}^\prime\) denotes a collection of
counterfactuals and \(\text{MMD}(\cdot)\) is the unbiased estimate of
the squared population maximum mean discrepancy, proposed by
\citet{gretton2012kernel}. The metric in Equation~\ref{eq-impl-div} is
equal to zero if and only if the two distributions are exactly the same,
\(\mathbf{X}^\prime=\mathbf{X}^+\).

For predictive performance, we use standard metrics, such as robust
accuracy estimated on adversarially perturbed data using FGSM
\citep{goodfellow2014explaining}.

We run experiments with three gradient-based generators: \emph{Generic}
of \citet{wachter2017counterfactual} as a simple baseline approach,
\emph{REVISE} \citep{joshi2019realistic} that aims to generate plausible
counterfactuals using a surrogate Variational Autoencoder (VAE), and
\emph{ECCo}---the generator of \citet{altmeyer2024faithful} without the
conformal prediction component---as a method that directly targets both
faithfulness and plausibility of the counterfactuals.

We make use of nine classification datasets common in the CE/AR
literature. Four of them are synthetic with two classes and different
characteristics: linearly separable clusters (\emph{LS}), overlapping
clusters (\emph{OL}), concentric circles (\emph{Circ}), and interlocking
moons (\emph{Moon}). They are generated using the library of
\citet{altmeyer2023explaining} and we present them in the supplementary
appendix. Next, we have four real-world binary tabular datasets:
\emph{Adult} (Census data) of \citet{becker1996adult2}, California
housing (\emph{CH}) of \citet{pace1997sparse}, Default of Credit Card
Clients (\emph{Cred}) of \citet{yeh2016default}, and Give Me Some Credit
(\emph{GMSC}) from \citet{kaggle2011give2}. Finally, for the convenience
of illustration, we use the 10-class \emph{MNIST}
\citep{lecun1998mnist}.

@@

Our main quantitative results for \emph{MLP} models are summarised in
Table~\ref{tbl-main}, which presents average outcomes along with
bootstrapped two standard errors. The following example motivates CT and
illustrates how to read Table~\ref{tbl-main}.

@@

\subsubsection{Synthetic Example: Prediction of Credit Card
Defaults.}\label{synthetic-example-prediction-of-credit-card-defaults.}

Figure~\ref{fig-poc} presents results for a linear classifier fitted to
\emph{LS} that complies with the data assumptions in Proposition
\ref{prp-mtblty}. The four panels show the outcomes for different
training procedures: in panels (a) and (c) we have trained the models
conventionally, while in panels (b) and (d) we have applied CT. For
illustrative purposes, suppose the first feature represents \emph{debt}
(mutable) and the second feature represents \emph{age} (immutable) of
loan applicants seeking counterfactual explanations for moving to the
target class: loan provided (orange).

@@

In panel (c) of Figure~\ref{fig-poc}, the CEs involve substantial
reductions in \emph{debt} for younger applicants. By comparison,
counterfactual paths are shorter on average in panel (d) where we have
protected the immutable \emph{age} as described in
Section~\ref{sec-constraints}. Due to the classifier's lower sensitivity
to \emph{age}, recommendations with respect to \emph{debt} are much more
homogenous and do not unfairly punish younger individuals. These CEs are
also plausible with respect to the mutable feature, despite requiring
smaller debt reductions on average, resulting in smaller costs to
individuals. This result is quantified for the non-linear case in column
three of Table~\ref{tbl-main}, which shows the \(\%\)-reduction in costs
averaged across valid counterfactuals. Once again, the impact of CT is
statistically significant and substantial (14\%). Thus, we consider the
model in panel (d) as the most explainable according to Def.
\ref{def-explainability}. Next, we present the results for all remaining
datasets.

@@

We find that CT generally leads to substantial and statistically
significant improvements in plausibility: average reductions in \(IP\)
range from around 7\% for \emph{MNIST} to almost 60\% for \emph{Circ};
for the real-world tabular datasets they are around 12\% for both
\emph{CH} and \emph{Cred} and almost 25\% for \emph{GMSC}; for
\emph{Adult} and \emph{OL} we find no significant impact of CT on
\(IP\). Reductions in \(IP^{*}\) are even more substantial and generally
statistically significant, although the average degree of uncertainty is
higher than for \(IP\): average reductions range from around 20\%
(\emph{Moons}) to almost 90\% (\emph{Circ}). The only negative findings
for \emph{OL} and \emph{MNIST} are statistically insignificant and, for
MNIST, do not align with qualitative findings, which are much more
plausible for CT (Figure~\ref{fig-mnist}).

@@

\caption{\label{fig-mnist}Sample explanations for \emph{MNIST} for BL
(top) and CT (bottom). First column is a random factual 0 (blue).
Columns 2 to 5 are corresponding \emph{ECCo} counterfactuals in target
classes 1 to 4. Columns 6 to 10 show integrated gradients averaged over
all test images in classes 5 to 9.}

@@

We also find that CT can reduce sensitivity to immutable, protected
features and thus lead to less costly counterfactual outcomes as shown
in Figure~\ref{fig-poc}. In column three of Table~\ref{tbl-main}, we
impose mutability constraints on selected features and compute the
reduction in average costs of CEs associated with CT compared to the
baseline: for synthetic datasets, we always protect the first feature;
for all real-world tabular datasets we could identify and protect an
\emph{age} variable; for \emph{MNIST}, we protect the five top and
bottom rows of digits. Reductions in costs are overwhelmingly positive
and significant of up to nearly 60\% for \emph{GMSC}. While the
estimated cost reductions for \emph{Adult} and \emph{MNIST} are not
significant, Figure~\ref{fig-mnist} (columns 6-10) demonstrates that CT
does have the expected effect: sensitivity to protected features as per
integrated gradients is drastically reduced; details of this experiment
are reported in the supplementary appendix. In the case of \emph{Cred},
average costs increase, likely because any potential benefits from
protecting the \emph{age} are outweighed by the increase in costs
required for greater plausibility.

@@

\caption{\label{tbl-main}Key performance metrics and bootstrapped
\emph{two} standard errors for all datasets. \textbf{Plausibility}
(columns 1-2): percentage reduction in implausibility for \(IP\) and
\(IP^*\), respectively; \textbf{Actionability} (column 3): percentage
reduction in costs with protected features. \textbf{Accuracy} (columns
4-7): test accuracies and robust accuracies (\(\text{Acc}^*\)) for CT
and the baseline (BL). Counterfactual outcomes in columns 1-3 are
aggregated across bootstrap samples and varying degrees of the energy
penalty \(\lambda_{\text{egy}}\) used for \emph{ECCo} at test time.
Standard errors for accuracy are bootstrapped from the test set.}

@@

\emph{CT increases the training times.} Like adversarial training, CT is
more resource-intensive than conventional regimes. Higher numbers of CEs
improve the quality of learned representations but they also increase
the number of computations. As our codebase is not performance
optimized, grids of 270 settings for the largest datasets in our
experiments took up to four hours using 34 2GB CPUs (see supplementary
appendix). Other than optimization, three factors mitigate this effect:
(1) CT yields itself to parallel execution; (2) it amortizes the cost of
CEs for the training samples; and (3) it can be used to fine-tune
conventionally-trained models.

We also highlight three important directions for future research.
Firstly, it is an interesting challenge to extend CT beyond
classification settings. Our formulation relies on the distinction
between non-target class(es) \(y^{-}\) and target class(es) \(y^{+}\) to
generate counterfactuals through Equation~\ref{eq-obj}. While \(y^{-}\)
and \(y^{+}\) can be arbitrarily defined, CT requires the output space
\(\mathcal{Y}\) to be discrete. Thus, it does not apply to ML tasks
where the change in outcome cannot be readily discretized. Focus on
classification is a common choice in research on CEs and AR. Other
settings have attracted some interest, e.g., regression
\citep{spooner2021counterfactual}, but there is little consensus how to
robustly extend the notion of CEs.\\
Secondly, CT is susceptible to training instabilities. This problem has
been recognized for JEMs \citep{grathwohl2020your} and even though we
depart from the SGLD sampling, we still encounter variability in
outcomes. CT is exposed to two potential sources of instabilities: (1)
the energy-based contrastive divergence term, \(\text{div}(\cdot)\), in
Equation~\ref{eq-div}, and (2) the underlying explainers. We find
several promising ways to mitigate this problem: regularizing energy
(\(\lambda_{\text{reg}}\)), generating sufficiently many counterfactuals
during each epoch, and including only mature counterfactuals in
\(\text{div}(\cdot)\).\\
Finally, we believe that it is possible to considerably improve
hyperparameter selection procedures, and thus performance. We have
relied exclusively on grid searches, but future research could benefit
from more sophisticated approaches.

To conclude, state-of-the-art machine learning models are prone to
learning complex representations that cannot be interpreted by humans.
Existing explainability solutions cannot guarantee that explanations
agree with the model's learned representation of data. As a step towards
addressing this challenge, we introduced counterfactual training, a
novel training regime that incentivizes highly-explainable models. Our
approach leads to explanations that are both plausible (compliant with
the underlying data-generating process) and actionable (compliant with
user-specified mutability constraints), and thus meaningful to their
recipients. Through extensive experiments, we demonstrated that CT
satisfies its objective while promoting robustness and preserving the
predictive performance of models. It can also be used to fine-tune
conventionally-trained models and achieve similar gains. Lastly, our
work highlights the value of simultaneously improving models and their
explanations.

### 2025-07-10 paper/sections/discussion.qmd
Trim to 7 pages

We also highlight several important directions for future research. Firstly, it is an interesting challenge to extend CT beyond classification settings. Our formulation relies on the distinction between non-target class(es) $y^{-}$ and target class(es) $y^{+}$ to generate counterfactuals through @eq-obj. While $y^{-}$ and $y^{+}$ can be arbitrarily defined, CT requires the output space $\mathcal{Y}$ to be discrete. Thus, it does not apply to ML tasks where the change in outcome cannot be readily discretized. Focus on classification is a common choice in research on CEs and AR. Other settings have attracted some interest, e.g., regression [@spooner2021counterfactual], but there is little consensus how to robustly extend the notion of CEs.  
Secondly, CT is susceptible to training instabilities. This problem has been recognized for JEMs [@grathwohl2020your] and even though we depart from the SGLD sampling, we still encounter variability in outcomes. CT is exposed to two potential sources of instabilities: (1) the energy-based contrastive divergence term, $\text{div}(\cdot)$, in @eq-div, and (2) the underlying explainers. We find several promising ways to mitigate this problem: regularizing energy ($\lambda_{\text{reg}}$), generating sufficiently many counterfactuals during each epoch, and including only mature counterfactuals in $\text{div}(\cdot)$.  
Finally, we believe that it is possible to substantially improve hyperparameter selection procedures, and thus performance. In this work, we have relied exclusively on grid searches for this task. Future work on CT could benefit from investigating more sophisticated approaches. Notably, our method is iterative which makes methods such as Bayesian or gradient-based optimization applicable [@bischl2023hyperparameter].

>>>>

*CT increases the training times.* Like adversarial training, CT is more resource-intensive than conventional regimes. Higher numbers of CEs improve the quality of learned representations but they also increase the number of computations. As our codebase is not performance optimized, grids of 270 settings for the largest datasets in our experiments took up to four hours using 34 2GB CPUs (see supplementary appendix). Other than optimization, three factors mitigate this effect: (1) CT yields itself to parallel execution; (2) it amortizes the cost of CEs for the training samples; and (3) it can be used to fine-tune conventionally-trained models.

We also highlight three important directions for future research. Firstly, it is an interesting challenge to extend CT beyond classification settings. Our formulation relies on the distinction between non-target class(es) $y^{-}$ and target class(es) $y^{+}$ to generate counterfactuals through @eq-obj. While $y^{-}$ and $y^{+}$ can be arbitrarily defined, CT requires the output space $\mathcal{Y}$ to be discrete. Thus, it does not apply to ML tasks where the change in outcome cannot be readily discretized. Focus on classification is a common choice in research on CEs and AR. Other settings have attracted some interest, e.g., regression [@spooner2021counterfactual], but there is little consensus how to robustly extend the notion of CEs.  
Secondly, CT is susceptible to training instabilities. This problem has been recognized for JEMs [@grathwohl2020your] and even though we depart from the SGLD sampling, we still encounter variability in outcomes. CT is exposed to two potential sources of instabilities: (1) the energy-based contrastive divergence term, $\text{div}(\cdot)$, in @eq-div, and (2) the underlying explainers. We find several promising ways to mitigate this problem: regularizing energy ($\lambda_{\text{reg}}$), generating sufficiently many counterfactuals during each epoch, and including only mature counterfactuals in $\text{div}(\cdot)$.  
Finally, we believe that it is possible to considerably improve hyperparameter selection procedures, and thus performance. We have relied exclusively on grid searches, but future research could benefit from more sophisticated approaches.

### 2025-07-10 paper/sections/method.qmd
Trim to 7 pages

`\noindent`{=latex} From a Bayesian perspective, setting $\mathbf{x}^+[d] := \mathbf{x}^\prime[d]$ can be understood as assuming a point mass prior for $p(\mathbf{x}^+)$ with respect to feature $d$. Intuitively, we think of this as ignoring implausibility costs with respect to immutable features, which effectively forces the model to instead seek plausibility with respect to the remaining features. This can be expected to result in lower overall sensitivity to immutable features, which we investigate empirically in @sec-experiments. Under certain conditions, this result holds theoretically:^[For the proof, see the supplementary appendix.]

>>>>

`\noindent`{=latex} From a Bayesian perspective, setting $\mathbf{x}^+[d] := \mathbf{x}^\prime[d]$ can be understood as assuming a point mass prior for $p(\mathbf{x}^+)$ wrt. feature $d$. Intuitively, we think of this as ignoring implausibility costs of immutable features, which effectively forces the model to instead seek plausibility through the remaining features. This can be expected to result in lower overall sensitivity to immutable features, which we investigate empirically in @sec-experiments. Under certain conditions, this result holds theoretically; for the proof, see the supplementary appendix:
